---
title: "Variation-in-selection.Rmd"
author: '"Yedra García"'
date: "2024-04-09"
output:
  pdf_document: default
  word_document: default
---

---
title: "Variation in patterns of selection and nonlinear selection"
author: "Yedra García"
date: "9 April 2026"
output: pdf_document
fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("nameofthepackage")
library(knitr)
library(dplyr)
library(reshape2)
```

## Spatiotemporal variation in phenotypic selection

Environmental variation is expected to generate variation in patterns of selection over time and space. For instance, individuals from different populations may experience varying selective environments and pressures, such as different pollinators, herbivores and predators. These fluctuating selective environments can lead to variation in individuals trait values and in the individuals fitness, ultimately altering the relationship between phenotypic traits and fitness (i.e. selection). The opportunity for selection, $I$ (variance in relative fitness $I = var(\frac{W}{\bar{W}})$) may thus also vary across different environments.

Similarly, the abiotic environment (nutrient availability, soil moisture, temperature) may also vary in space and time, even within the same population, potentially altering the strength and mode of selection. Thus, in order to understand how fluctuations in selection may influence biodiversity, we need to address the spatial and temporal scales at which selection varies (Albertsen et al. 2021). 

Typically, selection studies addressing spatiotemporal variation in selection have followed the regression approach by Lande and Arnold (1983) to estimate phenotypic linear selection gradients $\beta$ for each trait in each environment (e.g. population, year, experimental treatment) followed by analysis of covariance (ANCOVA) to test for differences in mean selection gradients $\beta$ among different environments.

Recent approaches, as the one proposed by Albertsen et al. (2021), now allow to comprehensively address variation in phenotypic selection over time and space by accounting for the uncertainty associated to any estimate of phenotypic selection and thus quantifying the magnitude of variation in selection.

Here, we will use a dataset from an selection study on different experimental populations (treatments) of the plant common morning glory, to address variation in phenotypic selection.

## Exercise 1: Selection on *Ipomoea purpurea* floral traits under water deficit

In this study by Garcia et al. (2023), we set up a common garden with the common morning glory $Ipomoea$ $purpurea$. We studied phenotypic selection in three different experimental populations or treatments consisting on 1) well-watered plants (watered every other day) exposed to open pollination and which we called $control$ treatment, 2) plants under water deficit named $drought$ treatment, and 3) well-watered plants with restricted access to pollinators, called $pollinator$ $restriction$ treatment. The motivation of this study was to address whether two selective agents on floral traits such as pollinators and abiotic resources (water availability) that are expected to change within a context of global change (pollinator decline and increasing droughts) may have an impact on phenotypic selection on floral traits related to plant pollination.

During the flowering season, we measured several phenotypic traits that function as floral signals and rewards to pollinators including flower size, nectar volume and concentration, and plant size (estimated from the diameter of the plant's main stem). We counted the number of fruits and estimated seed set (mean number of seeds per fruit multiplied by fruits) as a reproductive fitness measure.

\pagebreak

```{r}
ipo= read.csv("ipomoea_data.csv")
head(ipo)
ipo$Treatment= as.factor(ipo$Treatment)
ipo[1:5, 7:12]# the phenotypic traits 
```

#Phenotypic correlations between pairs of traits
```{r}
cor.ipo= cor(ipo[,c(7:10, 12)], use="pairwise")
signif(cor.ipo, 2)
```

#Explore the data in each experimental population (treatment)
```{r, fig.height=4, fig.width=14}
#In the control treatment (well-watered plants)
control= ipo %>%
  filter(Treatment %in% "C")

par(mfcol= c(1,4))
plot(control$mean_size, control$seedset)
plot(control$nectar_vl, control$seedset)
plot(control$nectar_conc, control$seedset)
plot(control$diameter, control$seedset)
```


```{r}
#Estimating relative fitness in the control treatment
control$relfit= control$seedset/mean(control$seedset, na.rm= TRUE)
```
Notice that we scale the traits for each population/treatment/year separately. In the same way as we do for trait standardization, we estimate the relative fitness $w = W/\bar{W}$ for each population, this is dividing individual fitness by its population's mean fitness.

```{r, fig.height=4, fig.width=14}
#In the drought treatment (plants in water deficit)
drought= ipo %>%
  filter(Treatment %in% "D")

par(mfcol= c(1,4))
plot(drought$mean_size, drought$seedset)
plot(drought$nectar_vl, drought$seedset)
plot(drought$nectar_conc, drought$seedset)
plot(drought$diameter, drought$seedset)
```


```{r}
#Estimating relative fitness in the drought (water deficit) treatment
drought$relfit= drought$seedset/mean(drought$seedset, na.rm= TRUE)
```

#In the pollinator restriction treatment (well-watered plants without pollinator access)
```{r, fig.height=4, fig.width=14}
#In the control treatment (well-watered plants)
pol.res= ipo %>%
  filter(Treatment %in% "PR")

par(mfcol= c(1,4))
plot(pol.res$mean_size, pol.res$seedset)
plot(pol.res$nectar_vl, pol.res$seedset)
plot(pol.res$nectar_conc, pol.res$seedset)
plot(pol.res$diameter, pol.res$seedset)
```


```{r}
#Estimating relative fitness in the pollination restriction treatment
pol.res$relfit= pol.res$seedset/mean(pol.res$seedset, na.rm= TRUE)
```

#Opportunity for selection (variance in relative fitness) in each treatment
```{r}
#In the control treatment
IC = var(control$relfit, na.rm= TRUE)
IC

#In the drought treatment
ID = var(drought$relfit, na.rm= TRUE)
ID

#In the pollinator restriction treatment
IPR = var(pol.res$relfit, na.rm= TRUE)
IPR

#Opportunity for selection is higher in the water deficit experimental population
```
#Multivariate variance-standardized selection gradients:

```{r}
#In the control treatment
m = lm(relfit~ scale(nectar_vl) + scale(nectar_conc)
+ scale(mean_size) + scale(diameter), data= control, na=na.exclude)
summary(m)$coef
```
We notice that there is linear selection to increase nectar volume (positive sign of the parameter's Estimate) in flowers in control conditions. This is, there is an increase of 9% (nectar vol Estimate= 0.09) in plant fitness (seed set) per standard deviation increase in nectar volume.
In addition, we detect selection to increase plant size (stem diameter), this is, larger plants produce more seeds. In this case seed set increases by 6% per standard deviation increase in plant size (stem diameter Estimate= 0.058)

#Multivariate mean-standardized selection gradients:

```{r}
#We center the traits to their mean (subtract the mean)
#and mean-scale the traits (divide by the mean)
control$nectar_vl_m = 
  (control$nectar_vl-mean(control$nectar_vl, na.rm=T))/mean(control$nectar_vl, na.rm=T)
control$nectar_conc_m = 
  (control$nectar_conc-mean(control$nectar_conc, na.rm=T))/mean(control$nectar_conc, na.rm=T)
control$mean_size_m = 
  (control$mean_size-mean(control$mean_size, na.rm=T))/mean(control$mean_size, na.rm=T)
control$diameter_m =
  (control$diameter-mean(control$diameter, na.rm = T))/mean(control$diameter, na.rm = T)
```

```{r}
#Then we estimate the mean-scaled multivariate linear selection gradients:
m.mean = lm(relfit~ nectar_vl_m + nectar_conc_m
+ mean_size_m + diameter_m, data= control, na=na.exclude)

summary(m.mean)$coef
```
We detect selection on the same traits, when we mean-scale the traits, but in this case the mean-scaled selection gradients, $\beta_\mu$, for nectar volume and stem diameter (0.294 vs 0.352) rank differently than variance-scaled gradients, $\beta_\sigma$ (0.098 vs 0.058)

#In the water deficit treatment:
```{r}
#Variance-scaled multivariate selection gradients: 
md = lm(relfit~ scale(nectar_vl) + scale(nectar_conc)
+ scale(mean_size) + scale(diameter), data= drought, na=na.exclude)

summary(md)$coef
```
In contrast to control plants, we observe selection favoring plants with lower nectar volumes (negative sign of the parameter's Estimate) suggesting a cost of nectar rewards in stressful conditions (water deficit). Specifically, fitness decreases by 13% (-0.133) per standard deviation increase in nectar volume. Interestingly, we find strong positive selection on floral size under water deficit (0.250), with an increase of 25% in fitness by a standard deviation increase in flower size.

#Mean-standardized selection gradients in water deficit:
```{r}
#First we center to the mean (subtract) and mean-scale (divide) the traits:
drought$nectar_vl_m =
  (drought$nectar_vl-mean(drought$nectar_vl, na.rm=T))/mean(drought$nectar_vl, na.rm=T)
drought$nectar_conc_m = 
  (drought$nectar_conc-mean(drought$nectar_conc, na.rm=T))/mean(drought$nectar_conc, na.rm=T)
drought$mean_size_m = 
  (drought$mean_size-mean(drought$mean_size, na.rm=T))/mean(drought$mean_size, na.rm=T)
drought$diameter_m =
  (drought$diameter-mean(drought$diameter, na.rm=T)) /mean(drought$diameter, na.rm = T)

#Then we estimate the multivariate linear selection gradients:
md.mean = lm(relfit~ nectar_vl_m + nectar_conc_m
+ mean_size_m + diameter_m, data= drought, na=na.exclude)

summary(md.mean)$coef
```
Again, we observed strong selection on flower size, which is stronger than selection on fitness as a trait($\beta_\mu = 1.549 > 1$).

As we have seen, the direction in patterns of selection on nectar volume differ between treatments, with positive directional selection in plants from control conditions contrasting with negative directional selection in plants under water deficit. Selection to increase flower size in water deficit also differs from what we observed in well-watered conditions (control and PR treatments), where we did not detect selection acting on flower size.

To estimate whether selection varies between treatments, we perform analysis of covariance (ANCOVA) in which we compare the means of the selection gradients $\beta$ between treatments. For this, we need to add the interaction terms between each trait and the variable that represents the different treatments (here "Treatment").
```{r}
#Estimate relative fitness within each treatment:
ipo.treat=ipo %>% 
  group_by(Treatment)%>%
  mutate("relfit"= seedset/mean(seedset, na.rm= TRUE))%>%
  ungroup()
```
Here, because of the unbalanced design (large difference in sample size between treatments, with larger size in the control treatment) we performed a paired ANCOVA between the control and the two manipulative treatments (i.e., water deficit and  and poll. restriction).

Control vs water deficit:
```{r}
#Fit the ANCOVA model to test for differences in selection between
#control and water deficit:
ipo.treat.cd = ipo.treat %>%
  filter(!Treatment %in% "PR") %>% #exclude pol.res treatment for comparison
  droplevels()

#Notice that the ANCOVA includes the interaction (*) between each trait and the treatment variable
m.ancova.cd= aov(relfit ~ scale(nectar_vl)+ scale(nectar_conc)
+ scale(mean_size)+ scale(diameter)+ scale(nectar_vl)*Treatment+
scale(nectar_conc)*Treatment+ scale(mean_size)*Treatment+
  scale(diameter)*Treatment, na= na.exclude, data = ipo.treat.cd)
summary(m.ancova.cd)
```

The ANCOVA indicates that phenotypic selection on nectar volume and floral size differs between the control and the water deficit treatment.

\pagebreak

Control vs poll. restriction:
```{r}
ipo.treat.cpr = ipo.treat %>%
  filter(!Treatment %in% "D") %>% #exclude water deficit treatment for #comparison
  droplevels()

m.ancova.cpr= aov(relfit ~ scale(nectar_vl)+ scale(nectar_conc)+
scale(mean_size)+ scale(diameter)+ scale(nectar_vl)*Treatment+ 
  scale(nectar_conc)*Treatment+ scale(mean_size)*Treatment+ scale(diameter)*Treatment,
na= na.exclude, data = ipo.treat.cpr)

summary(m.ancova.cpr)
```
In contrast, selection did not differ between well-watered plants with varying access to pollinators (i.e control vs. pollinator restriction treatment)

#ANCOVA on mean-standardized multivariate selection gradients:
```{r}
#Another way to center and scale the traits by their mean within each treatment
ipo.treat= ipo.treat %>% 
  group_by(Treatment)%>%
  mutate(across(c(nectar_vl, nectar_conc, mean_size, diameter),
                ~(.x-mean(.x, na.rm = TRUE))/mean(.x, na.rm = TRUE),
                .names= "{col}_m"))%>%
  ungroup()

#Fit the ANCOVA model to test for differences in selection between control and water deficit:
ipo.treat.cd= ipo.treat %>%
  filter(!Treatment %in% "PR") %>% #exclude pol.res treatment for
  #comparison
  droplevels()
  
m.ancova.cdm= aov(relfit ~ nectar_vl_m+ nectar_conc_m+ mean_size_m+
diameter_m+ nectar_vl_m*Treatment+ nectar_conc_m*Treatment+ 
  mean_size_m*Treatment+ diameter_m*Treatment,
na= na.exclude, data = ipo.treat.cd)

summary(m.ancova.cdm)
```
The ANCOVA approach indicates variation in phenotypic selection between the control
and water deficit treatment. To evaluate the magnitude of variation 
in phenotypic selection gradients between treatments, we perform the Error-corrected standard deviation of the selection gradients (Albertsen et al. 2021).

#Error-corrected variance of the mean-standardized phenotypic selection gradients:

To further evaluate variation in phenotypic selection between treatments, we follow
the approach proposed by Albertsen et al (2021) that estimates the mean dispersion of selection gradients.

We start by fitting the multivariate selection model on our set of traits on each treatment separately. We will focus on the comparison between the control and the drought treatment for which we have seen that the ANCOVA test supports variation in patterns of selection.

This time, we will fit the multivariate selection models just on the mean-centered traits (subtracting the trait mean), and later we will estimate the mean-scaled gradients by multiplying the raw model's estimates by the trait means. 

```{r}
#Control treatment. 
#Mean-center the traits (subtract the mean trait value)
control= control %>%
  mutate(
    nectar_vl_mean_cent   = nectar_vl - mean(nectar_vl, na.rm = TRUE),
    nectar_conc_mean_cent = nectar_conc - mean(nectar_conc, na.rm = TRUE),
    mean_size_mean_cent   = mean_size - mean(mean_size, na.rm = TRUE),
    diameter_mean_cent    = diameter - mean(diameter, na.rm = TRUE))

```

Fit the multivariate regression (selection) model on the mean-centered traits
```{r}
control_mean_mod= lm(relfit~ nectar_vl_mean_cent + nectar_conc_mean_cent
+ mean_size_mean_cent + diameter_mean_cent, data= control, na=na.exclude)
```

Then we calculate the mean and standard variation for each trait 
```{r}
control_mean_traits= control %>% 
  summarise_at(c("nectar_vl",
                       "nectar_conc",
                       "mean_size",
                       "diameter"), mean,
               na.rm=TRUE)
```

Reshape the format with melt function
```{r, message= FALSE}
control_mean_traits= melt(control_mean_traits, value.name = "Trait_Mean", variable.name = "Trait")
control_mean_traits
```


```{r}
control_sd_traits= control %>% 
  summarise_at(c("nectar_vl",
                       "nectar_conc",
                       "mean_size",
                       "diameter"), sd,
               na.rm=TRUE)
control_sd_traits
```

```{r, message= FALSE}
control_sd_traits= melt(control_sd_traits, value.name = "Trait_SD", variable.name = "Trait")
control_sd_traits
```


```{r}
control_summary_traits= left_join(control_mean_traits,
                                                control_sd_traits,
                                                by = c("Trait" = "Trait"))
```

Round to four decimals
```{r}
control_summary_traits= control_summary_traits %>% 
  mutate_if(is.numeric, round, digits = 4)

control_summary_traits
```
Create a dataframe with the model's information (estimate for each trait, SE and p value).

To get the model's estimates for each trait, we just extract the model coefficients.
To estimate the standard error (SE) of the model's estimates (Estimate) for each trait, we calculate the standard errors as the square root (sqrt) of the variance, which is the diagonal of the variance-covariance (vcov(model)) matrix of the model.

We only obtain the estimates, SE and p. values, for the traits (model's coefficients[2:5]) and exclude the intercept (model's coefficient[1])

```{r}
control_betas=
  data.frame(Trait = c("nectar_vl",
                       "nectar_conc",
                       "mean_size",
                       "diameter"),
             Estimate = c(control_mean_mod$coefficients[2],
                          control_mean_mod$coefficients[3],
                          control_mean_mod$coefficients[4],
                          control_mean_mod$coefficients[5]),
             SE_Model = c(sqrt(diag(vcov(control_mean_mod))[2]),
                          sqrt(diag(vcov(control_mean_mod))[3]),
                          sqrt(diag(vcov(control_mean_mod))[4]),
                          sqrt(diag(vcov(control_mean_mod))[5])),
             p = c(summary(control_mean_mod)$coefficients[2, "Pr(>|t|)"],
                   summary(control_mean_mod)$coefficients[3, "Pr(>|t|)"],
                   summary(control_mean_mod)$coefficients[4, "Pr(>|t|)"],
                   summary(control_mean_mod)$coefficients[5, "Pr(>|t|)"]))
control_betas
```

Round to four decimals
```{r}
control_betas= control_betas %>% 
  mutate_if(is.numeric, round, digits = 4)
control_betas
```

```{r}
control_gradients= left_join(control_betas, control_summary_traits,
                                                     by = c("Trait" = "Trait"))
control_gradients
```

#Estimate mean-scaled selection gradients (and their SE) for each trait
As previously mentioned, we multiply the model raw estimates for each trait (Estimate) by the trait mean in order to estimate the mean-scaled selection gradients.
```{r}
control_gradients=  control_gradients %>% 
  mutate(Mean_Selection_Gradient = round(Estimate* Trait_Mean, digits = 4)) %>% 
  mutate(Mean_SE_Selection_Gradient = round(SE_Model * Trait_Mean, digits = 4)) %>% 
  mutate(Treatment = "control")

control_gradients
```

Now we follow the same steps for the water deficit treatment:
```{r}
#Drought treatment. 
#Mean-center the traits (subtract the mean trait value)
drought= drought %>%
  mutate(
    nectar_vl_mean_cent   = nectar_vl - mean(nectar_vl, na.rm = TRUE),
    nectar_conc_mean_cent = nectar_conc - mean(nectar_conc, na.rm = TRUE),
    mean_size_mean_cent   = mean_size - mean(mean_size, na.rm = TRUE),
    diameter_mean_cent    = diameter - mean(diameter, na.rm = TRUE))
```

Fit the multivariate selection models on the mean-centered traits
```{r}
drought_mean_mod= lm(relfit~ nectar_vl_mean_cent + nectar_conc_mean_cent
+ mean_size_mean_cent + diameter_mean_cent, data= drought, na=na.exclude)
  
```

Then we calculate the mean and standard variation for each trait 
```{r}
#drought treatment
drought_mean_traits= drought %>% 
  summarise_at(c("nectar_vl",
                       "nectar_conc",
                       "mean_size",
                       "diameter"), mean,
               na.rm=TRUE)

```

```{r, message= FALSE}
drought_mean_traits= melt(drought_mean_traits, value.name = "Trait_Mean", variable.name = "Trait")

drought_mean_traits
```


```{r, message= FALSE}
drought_sd_traits= drought %>% 
  summarise_at(c("nectar_vl",
                       "nectar_conc",
                       "mean_size",
                       "diameter"), sd,
               na.rm=TRUE)
drought_sd_traits
```


```{r, message= FALSE}
drought_sd_traits= melt(drought_sd_traits, value.name = "Trait_SD", variable.name = "Trait")
drought_sd_traits
```


```{r}
drought_summary_traits= left_join(drought_mean_traits,
                                                drought_sd_traits,
                                                by = c("Trait" = "Trait"))
```

Round to four decimals
```{r}
drought_summary_traits= drought_summary_traits %>% 
  mutate_if(is.numeric, round, digits = 4)

drought_summary_traits
```

Create a dataframe with the model information (estimate for each trait, SE and p value)
```{r}
drought_betas=
  data.frame(Trait = c("nectar_vl",
                       "nectar_conc",
                       "mean_size",
                       "diameter"),
             Estimate = c(drought_mean_mod$coefficients[2],
                          drought_mean_mod$coefficients[3],
                          drought_mean_mod$coefficients[4],
                          drought_mean_mod$coefficients[5]),
             SE_Model = c(sqrt(diag(vcov(drought_mean_mod))[2]),
                          sqrt(diag(vcov(drought_mean_mod))[3]),
                          sqrt(diag(vcov(drought_mean_mod))[4]),
                          sqrt(diag(vcov(drought_mean_mod))[5])),
             p = c(summary(drought_mean_mod)$coefficients[2, "Pr(>|t|)"],
                   summary(drought_mean_mod)$coefficients[3, "Pr(>|t|)"],
                   summary(drought_mean_mod)$coefficients[4, "Pr(>|t|)"],
                   summary(drought_mean_mod)$coefficients[5, "Pr(>|t|)"]))

```

Round to four decimals
```{r}
drought_betas= drought_betas %>% 
  mutate_if(is.numeric, round, digits = 4)
drought_betas
```


```{r}
drought_gradients= left_join(drought_betas, drought_summary_traits,
                                                     by = c("Trait" = "Trait"))
```

#Estimate the mean-scaled selection gradients in the drought treatment:
```{r}
drought_gradients=  drought_gradients %>% 
  mutate(Mean_Selection_Gradient = round(Estimate* Trait_Mean, digits = 4)) %>% 
  mutate(Mean_SE_Selection_Gradient = round(SE_Model * Trait_Mean, digits = 4)) %>% 
  mutate(Treatment = "drought")

drought_gradients
```


```{r}
data_gradients= rbind(control_gradients, drought_gradients)
data_gradients
```

##Estimate the error-corrected SD of the Mean selection gradients:
```{r, warning=FALSE}
data_gradients_mean_variance= data_gradients %>% 
  group_by(Trait) %>% 
  summarise(Variance_Mean_Selection_Gradient = var(Mean_Selection_Gradient),
            Mean_Mean_SE_Selection_Gradient = mean(Mean_SE_Selection_Gradient^2),
            Difference = Variance_Mean_Selection_Gradient - Mean_Mean_SE_Selection_Gradient,
Variance_Corrected_Mean_Selection_Gradient = 
sqrt(Variance_Mean_Selection_Gradient - Mean_Mean_SE_Selection_Gradient)) %>% 
  mutate_if(is.numeric, round, digits = 4)
data_gradients_mean_variance
```
We see that in two  out of the four traits (floral size and nectar volume) there is variation in phenotypic selection between the two treatments (control vs drought).
The magnitude of the variation in selection on floral size was three times (1.205) the magnitude of variation on nectar volume (0.396). In contrast, there is no detectable variation in selection on plant size and nectar concentration beyond sampling error.
\pagebreak


## Nonlinear selection

#Stabilizing and disruptive selection

Up to now we have seen different examples of linear, also known as directional, phenotypic selection. This mode of selection acts by changing the mean trait values in the population (e.g. plants with larger flowers or lower nectar volumes are selected).
However, there are other modes of selection, where selection acts by changing the phenotypic variance in the population rather than the mean. For instance, selection may favor those individuals with trait values closer to the average phenotype (they will have higher fitness), while individuals with trait values close to the extremes of the distribution will show lower fitness values. This mode of selection is known as $stabilizing$ $selection$ and it acts decreasing the phenotypic variance in the population. 
In contrast to stabilizing selection, selection may favor those individuals with extreme trait values (close to the tails of the trait distribution) acting against average individuals. This case of nonlinear selection is called $disruptive$ $selection$ and increases the phenotypic variance in the population.

In both cases, we will need curve-fitting approaches to describe the selection surface (the relationship between the individuals phenotypic trait values and fitness). The regression approach by Lande and Arnold (1983) also allow us to quantify nonlinear phenotypic selection gradients by applying second‐order polynomial regression. For this, we add a quadratic term to the regression model of relative fitness on phenotype, which can be written as:

$w = \alpha + \sum_{i} z_i\beta_i + 1/2 \sum_{i} z_i^2\gamma_i+ \epsilon_i$

where $w$ is the relative fitness of the individual, $\alpha$ is an intercept, $\beta$'s are partial regression coefficients (linear regression terms) and $\gamma$'s are the quadratic coefficients of the regression model of relative fitness ($w$) on the individual's trait $zi$ values.
The 1/2 factor is used to make the quadratic terms equivalent to second derivatives (Arnold 2003).

The quadratic regression terms $\gamma_i$'s are used to measure nonlinear selection.$\gamma_i$'s <0 indicate stabilizing selection while $\gamma_i$'s >0 indicate disruptive selection. It is important to notice that to obtain the nonlinear selection gradients, we need to double the resulting quadratic regression coefficients, $\gamma_i$, as well as their standard error (Stinchcombe et al. 2008)

#Correlational selection

Another example of nonlinear selection is correlational selection, in which individuals with specific combinations of trait values  show higher fitness (Phillips and Arnold 1989). This mode of selection is known as correlational selection and implies that selection on one trait depends on the value of another trait. 
Once again, we can use Lande and Arnold's regression approach to quantify correlational selection. To do that we add the cross-product terms $\gamma_{i,j}$ for each pair of traits to the regression model with the linear and the quadratic regression terms. For two traits $z_i$ and $z_j$ the model can be written as:

$w = \alpha + z_i\beta_i+ z_j\beta_j+ 1/2 z_i^2\gamma_{i,i}+ 1/2 z_j^2\gamma_{j,j}+ z_iz_j\gamma_{i,j}+ \epsilon$

where $\gamma_{i,j}$ denotes the correlational selection gradient for traits $z_i$ and $z_j$

To explore nonlinear selection, we will use a dataset from a study by Campbell et al. (2022) on $Ipomopsis$

## Exercise 2: Selection by pollinators and seed predators on floral traits in two *Ipomopsis* species.

In this work by Dianne Campbell et al. (Campbell et al. 2022), they investigated linear and nonlinear phenotypic selection acting on a set of floral traits, including floral size-related traits, nectar, color and scent, in two species of $Ipomopsis$ (pollinated by hummingbirds and hawk moths) and their hybrids during two different stages of the plant's life cycle which are: 1) pollination (i.e. seeds initiated) and 2) seed predation (proportion of fruits escaping fly predation).

Previous work on this system has shown that certain traits are under directional selection by pollinators and seed predators. Based on this knowledge, here they tested several predictions in which correlational selection may arise as the effect of combining directional selection by both pollinators and seed predators.

Here, we will focus on $I.$ $aggregata$ and three traits (floral width, color and nectar production) for which sample sizes are large enough to test for linear, quadratic and correlational selection through  multivariate regression following Lande and Arnold's approach.
```{r}
camp22= read.csv("Campbell_etal_22.csv")
names(camp22)
camp22$Site= as.factor(camp22$Site)
camp22$Type= as.factor(camp22$Type)
```

#Filter to include only I. aggregata sites
```{r}
aggre= camp22 %>%
  filter(Site== "agg") %>% 
dplyr::select(c(1:6, 11, 12))%>% #Filter to include petal width, color #and nectar
droplevels()
```
Explore the data: at pollination stage (seeds initiated)
```{r, fig.height=4, fig.width=10}
par(mfrow= c(1,3))
plot(aggre$width, aggre$seedsinit)
plot(aggre$color, aggre$seedsinit)
plot(aggre$nectar, aggre$seedsinit)
```

Explore the data: at predation stage (seeds escaping fly predation)
```{r, fig.height=4, fig.width=10}
par(mfrow= c(1,3))
plot(aggre$width, aggre$flyescape)
plot(aggre$color, aggre$flyescape)
plot(aggre$nectar, aggre$flyescape)
```
Estimate relative fitness at pollination and seed predation stages:
```{r}
aggre$relfit.pol = aggre$seedsinit/mean(aggre$seedsinit, na.rm = T)
aggre$relfit.pred = aggre$flyescape/mean(aggre$flyescape, na.rm = T)
```
#Variance-scaled univariate linear selection gradients at the pollination stage
```{r}
summary(lm(relfit.pol~ scale(width), data= aggre, na= na.exclude))$coef
summary(lm(relfit.pol~ scale(color), data= aggre, na= na.exclude))$coef
summary(lm(relfit.pol~ scale(nectar), data=aggre, na= na.exclude))$coef
```
We observe positive directional selection to increase flower width and negative directional selection to decrease petal color (selection for pale-colored flowers)

Now when we accounting for selection acting on other traits:
#Variance-scaled multivariate linear selection gradients at the pollination stage
```{r}
m = lm(relfit.pol~ scale(width) + scale(color)
+ scale(nectar), data= aggre, na= na.exclude)
summary(m)$coef
```
We still detect linear selection through seeds initiated (i.e. pollination stage) acting on flower width ($\beta_\sigma$= 0.48) and petal color ($\beta_\sigma$= -0.30)

#Mean-scaled multivariate linear selection gradients at pollination stage
```{r}
#Mean-scale the traits
aggre= aggre %>% 
mutate(across(c(width, color, nectar),~ (.x-mean(.x, na.rm = T))/mean(.x, na.rm = T),
.names= "{col}_m"))


m = lm(relfit.pol~ width_m + color_m+ nectar_m, na= na.exclude, data= aggre)
summary(m)$coef
```
Mean-standardized multivariate selection gradients also show strong directional selection on flower width (positive) and floral color (negative), with both $\beta_\mu$'s >1 (3.86 and -1.25) which are stronger than selection on fitness as a trait.

#Estimating multivariate nonlinear selection gradients (quadratic and interaction terms):
We fit one model with:
Linear terms: linear selection
Quadratic terms: stabilizing and disruptive selection
Interaction terms for each pair of traits: correlational selection

To add the quadratic terms to the model, we need to square the traits values:
```{r}
#We first variance-scale the traits (here with "scale" function= variance-scaled traits)
#and then we square (^2) the traits:

aggre= aggre %>% 
  mutate(across(c(width, color, nectar),  ~(scale(.)), .names= "{col}_s")) %>%
mutate(across(c(width_s, color_s, nectar_s),  ~ .x^2, .names= "{col}q"))
```

#Variance-scale multivariate nonlinear selection gradrients
```{r}
m = lm(relfit.pol~ width_s + color_s+ nectar_s+
         width_sq + color_sq+ nectar_sq+
  width_s*color_s+ width_s*nectar_s+  color_s*nectar_s, 
  na= na.exclude, data= aggre)
summary(m)$coef

```
There is Disruptive selection on floral color (color_sq$coeff= 0.27 >0)
To estimate the quadratic selection gradient $\gamma_i$ we double the quadratic regression coefficient and the standard error for flower color:

```{r}
gamma_color =summary(m)$coef[6, 1]*2
SE= 0.091*2

gamma_color
#gamma for floral color= 0.55 (0.18)
```

The model also indicates negative correlational selection ($\gamma_i,_j$= -0.512) acting on flower width and color. This, selection favors plants with wide flowers when flowers are pale in color.

#Mean-scaled multivariate nonlinear selection gradients:
```{r}
#Square the mean-scaled traits
aggre= aggre %>%
mutate(across(c(width_m, color_m, nectar_m),  ~ .x^2, .names= "{col}q"))

m = lm(relfit.pol~ width_m + color_m+ nectar_m+ width_mq +
         color_mq+ nectar_mq+ width_m*color_m+ width_m*nectar_m+
         color_m*nectar_m, na= na.exclude, data= aggre)
summary(m)$coef
```

